---
output: pdf_document
---

\newpage

# Pre-processing data outside R

R is good for many things but not everything.
As mentioned in the introduction, there are various ways to
preprocess large files outside R to make them easier to handle.
Here we will explore some of the options.

## Split the file with Unix tools

The unix utility **split** can be used to split large files up
by size or number of lines. The following bash commands will split the
5.6 GB file, downloaded and unzipped in the first chapter, into chunks of 100 MB
each:^['Bash commands'
refer to computer code written in the Bash language. Bash is the default
language used by Linux and Macs for most internal system administration
functions. In Macs, you can open this by typing 'Apple key'-T. In
Windows, installing [cygwin](https://www.cygwin.com/) and launching it
will provide access to this functionality. Note: you must start from the correct
*working directory*. Use `pwd` in Bash or `setwd()` in R to check this.]

```
cd npidata
split -b100m npidata_20050523-20150809.csv
```

Assuming there is sufficent
disk space, there output of the above operation should be several 100 Mb text
files which are more manageable. These are named `aa`, `ab` etc.
A sample from the results of this operation can be found in the
`sample-data` folder after the following commands.

```
split -l 10 aa mini # further split chunk 'aa' into 10 lines
cp miniaa ../sample-data # copy the first into 'sample-data'
```

Now the file is much smaller and easy to read: finally we can
read (part of) a 5.6 GB text file into R!

```{r}
npi <- read.csv("sample-data/miniaa")
dim(npi)
head(npi[c(1, 37)], 3)
```

## Filtering with csvkit

[csvkit](https://csvkit.readthedocs.org/en/latest/) is a command-line program
for handling large .csv files, without having to read them all into RAM...

Using the npi data, the following
[example](https://opendata.stackexchange.com/questions/1256/how-can-i-work-with-a-4gb-csv-file)
ilustrates how csvkit can be used to extract useful information from
bulky .csv files before loading the results into R.

```{r}
# ...
```

## Preprocessing with the LaF package




