---
output: pdf_document
---

\chapter{The RCpp package}

# Introduction

Sometimes R is just slow. You've tried every trick you know, and your code is still crawling along. At this point you may need to rewrite key parts of your code in C/C++. You don't have to resort to external packages to call C/Fortran routines. We can just use the `.Call` function; it's just an incredibly painful and error prone experience. However there is a better way, the Rcpp\sidenote{\url{http://www.rcpp.org}} package. This is now one of the most popular packages on CRAN.  Rcpp provides a clean and friendly API\sidenote{\textbf{A}pplication \textbf{P}rogram \textbf{I}nterface is an exposed  set of routines, protocols, and/or tools for building software applications.} that lets you write high-performance code, while at the same time keeping you safe from R's tricky C API. The typical bottlenecks that C/C++ can address are loops and recursive functions.

In this chapter, C and C++ code are largely interchangeable, so when you see 'C code', it can usually be included in a `.cpp` file\marginnote{`.cpp` is the default file extension for C++ scripts.}. In general this isn't true. See for example the Stackoverflow question, \url{http://programmers.stackexchange.com/q/16390/14846} for an overview.

C++ is a separate programming language, so this chapter just provides the bare minimum to get you started. This chapter's goal is to provide a flavour of what's possible.

\subsection*{Pre-requisites}

To use\sidenote{To use means being able to write and compile functions. If you distribute code in a package, then this isn't an issue. For example, the \texttt{ggplot2} package uses Rcpp.} the package you need a working C++ compiler. 

 * Linux: A compiler should already be installed. Otherwise install `r-base` it a compiler will be installed as a dependency.
 * Macs: Install `Xcode`.
 * Windows: Install Rtools\sidenote{\url{http://cran.r-project.org/bin/windows/}}. Make sure you select the version that corresponds to your version of R.

\noindent The code in this chapter was generated using version `r packageDescription("Rcpp")$Version` of Rcpp. You can install Rcpp from CRAN in the usual way

```{r eval=FALSE}
install.packages("Rcpp")
```

\noindent The associated CRAN\sidenote{\url{https://cran.r-project.org/web/packages/Rcpp/}} page has numerous vignettes that are worth reading\sidenote{You can get an idea of the popularity of the package by looking at the `Reverse Imports` section.}.

To check that you have everything needed for this chapter, run the following piece of code from the course R package

```{r cache=TRUE}
library("r4bd")
test_rcpp()
```

# A simple C++ function

A C/C++ function is similar to an R function: you pass a set of inputs to function, some code is run, a single object is returned. However there are some key differences.

1. In the C/C++ function, each line must be terminated with `;` In R, we use `;` only when we have multiple statements on the same line.
2. We must declare object types in the C/C++ version. In particular we need to declare the types of the function arguments, return value and any intermediate objects we create. 
3. The function must have an explicit `return` statement. Similar to R, there can be multiple returns, but the function will terminate when it hits it's first `return` statement.
4. You do not use assignment when creating a function.
5. Object assignment must use `=` sign. The `<-` operator isn't valid\sidenote{Yet another reason to use \texttt{=} when writing R code}.
6. One line comments can be created using `//`. Multi-line comments are created using `/*...*/`

\noindent We want to create a function that adds two numbers together. In R this would be a simple one line affair:

```{r} 
add_r = function(x, y) x + y
```

\noindent In C++ it is a bit more long winded

```{Rcpp eval=FALSE}
/* Return type double
 * Two arguments, also doubles
 */
double add_c(double x, double y) { 
    double value = x + y;
    return value;
  }
```

\noindent If we were writing C++ code, we would also need another function called `main`. We would then compile the code to obtain an executable that is run. The executable is platform dependent. The beauty of using Rcpp is that it makes it very easy to call C++ functions from R and the user doesn't have to worry about the platform, or compilers or the R/C interface.

## The cppFunction command
 
We load the Rcpp package using the usual `library` function call
 
```{r message=FALSE}
library("Rcpp") 
```
 
\noindent Then we simply pass the C++ function as a (string) argument to `cppFunction`:

```{r tidy=FALSE}
cppFunction('
  double add_c(double x, double y){
    double value = x + y;
    return value;
  }
')
```

\noindent and Rcpp will magically compile the C++ code and construct a function that bridges the gap between R and C++. After running the above code, we now have access to the `add_c` function

```{r}
add_c
```

\noindent We can call the `add_c` function in the usual way

```{r}
add_c(1, 2)
```

\noindent and we don't have to worry about compilers. It has all been taken care of. Also,  if you include this function in a package, users don't have to worry about any of the Rcpp magic. It just works.

## C/C++ data types

The most basic type of variable is an integer, `int`. An `int` variable can store a value in the range $-32768$ to $+32767$\sidenote{In C, we can also define unsigned \texttt{int}. Then the range goes from $0$ to $65,535$. There are also `long int` data types, which range from $0$ to $2^{31}-1$.}. To store floating point numbers, there are single precision numbers, `float` and double precision numbers, `double`. A `double` takes twice as much memory as a `float`.  For __single__ characters, we use the `char` data type.

\begin{table}[t]
  \centering
  \begin{tabular}{@{}ll@{}}
  \toprule
  Type & Description\\
  \midrule
  \texttt{char} & A single character.\\
  \texttt{int} & An integer.\\
  \texttt{float} & A single precision floating point number.\\
  \texttt{double} & A double-precision floating point number.\\
  \texttt{void} & A valueless quantity.\\
  \bottomrule
  \end{tabular}
  \vspace{0.2cm}
  \caption{Overview of key C/C++ object types.}
\end{table}

A pointer object is a variable that points to an area of memory that has been given a name. Pointers are a very powerful, but primitive facility contained in the C language. They are very useful since rather than passing large objects around, we pass a pointer to the memory location; rather than pass the house, we just give the address. We won't use pointers in this chapter, but mention them for completeness.

# The sourceCpp function

The `cppFunction` is great for getting small examples up and running. But it is better practice to put your C++ code in a separate file (with file extension `cpp`) and use the function call `sourceCpp("path/to/file.cpp")` to compile them. However we need to include a few headers at the top of the file. The first line we add gives us access to the Rcpp functions. The file `Rcpp.h` contains a list of function and class definitions supplied by Rcpp\sidenote{This file will be located where Rcpp is installed. Alternatively, you can view it online at \url{https://github.com/RcppCore/Rcpp}}. The `include` statement adds the definitions to the top of your code

```{Rcpp eval=FALSE}
#include <Rcpp.h>
```

\noindent To access the Rcpp functions we would have to type `Rcpp::function_1`. To avoid typing `Rcpp::`, we use the namespace facility

```{Rcpp eval=FALSE}
using namespace Rcpp;
```

\noindent Now we can just type `function_1`\sidenote{This is the same concept that R uses for managing function name collisions when loading packages.}. Above each function we want to export/use in R, we add the tag\sidenote{Similar to an Roxygen2 export tag.}

```{Rcpp eval=FALSE}
// [[Rcpp::export]]
```

\noindent This would give the complete file

```{Rcpp}
#include <Rcpp.h>
using namespace Rcpp;

// [[Rcpp::export]]
double add_c(double x, double y){
  double value = x + y;
  return value;
}
```

\noindent There are two main benefits with putting your C++ functions in separate files. First, we have the benefit of syntax highlighting (RStudio has great support for C++ editing). Second, it's easier to make syntax errors when the switching between R and C++ in the same file. To save space, we we'll omit the headers for the remainder of the chapter.

# Vectors and loops

Let's now consider a slightly more complicated example. Here we want to write our own function that calculates the mean. This is just an illustrative example: R's version is much better and more robust to scale differences in our data. For comparison, let's create a corresponding R function. The function takes a single vector `x` as input, and returns the mean value, `m`:

```{r}
mean_r = function(x) {
  n = length(x)
  m = 0
  for(i in seq_along(x)) 
    m = m + x[i]/n
  m
}
```

\noindent This is a very bad R function. We should just use the base function `mean` for real world applications. However the purpose of `mean_r` is to provide a comparison for the C++ version, which we will write in a similar way.

In this example, we will let Rcpp smooth the interface between C++ and R by using the `NumericVector` data type. This Rcpp data type mirrors the R vector object type. Other common classes are: `IntegerVector`, `CharacterVector`, and `LogicalVector`. 

In the C++ version of the mean function, we specify the arguments types: `x` (`NumericVector`) and the return value (`double`). The C++ version of the `mean` function is a few lines longer. Almost always, the corresponding C++ version will be, possibly much, longer. 

```{Rcpp eval=FALSE}
double mean_c(NumericVector x){
  int i;
  int n = x.size();
  double mean = 0;
  
  for(i=0; i<n; i++) {
    mean = mean + x[i]/n;
  }
  return mean;
}
``` 

\noindent To use the C++ function, we need to source the file (remember to put the necessary headers in).

```{r}
sourceCpp("src/mean_c.cpp")
```

\noindent Although the C++ version is similar, there are a few crucial differences.

1. We use the `.size()` method to find the length of `x`
1. The `for` loop has a more complicated syntax.
    ```{Rcpp eval=FALSE}
    for (variable initialization; condition; variable update ) {
       // Code to execute
    }
    ```
1. C++ provides operators to modify variables in place. So `i++` increases the value of `i` by `1`. Similarly, we could rewrite part of the loop as
    ```{Rcpp eval=FALSE}
    mean += x[i]/n;
    ```
   The above code adds `x[i]/n` to the value of `mean`. Other similar operators are `-=`, `*=`, `/=` and `i--`.
1. A C++ vector starts at `0` **not** `1`

To compare the C++ and R functions, we use the `microbenchmark` package.

```{r}
library("microbenchmark")
```

\noindent This package is useful for comparing functions that run quickly. It serves as a more accurate replacement to `system.time(replicate(1000, expr))`. We will include the base R `mean` function in the comparison. We generate some normal random numbers for the comparison

```{r}
x = rnorm(1e4)
```

\noindent Then call the `microbenchmark` function.

```{r cache=TRUE}
z = microbenchmark(
  mean(x),
  mean_r(x),
  mean_c(x)
)
```

\noindent The results are easily compared using the `boxplot` method

```{r fig.width=5, fig.height=3, echo=2, fig.cap="Comparison of mean functions.", echo=2, cache=TRUE, fig.pos='t'}
par(mar=c(3,3,2,1), mgp=c(2,0.4,0), tck=-.01, cex.axis=0.9, las=1)
boxplot(z, ylab="Log (time)")
grid()
```

\noindent In this simple example, the C++ variant is around $100$ times faster than the corresponding R version.

\subsection*{Exercise}

Consider the following piece of code
```{Rcpp eval=FALSE}
double test1() {
  double a = 1.0 / 81;
  double b = 0;
  for (int i = 0; i < 729; ++ i)
    b = b + a;
  return b;
}
```
\begin{enumerate}
\item Save the function `test1` in a separate file. Make sure it works.
\item Write a similar function in R and compare the speed of the C++ and R versions.
\item Create a function called `test2` where the `double` variables have been replaced by `float`. Do you still get the correct answer?
\item Change `b = b + a` to `b += a` to make you code more `C` like. 
\item (Hard) What's the difference between `i++` and `++i`?
\end{enumerate}

## Matrices

Each vector type has a corresponding matrix equivalent: `NumericMatrix`, `IntegerMatrix`, `CharacterMatrix` and `LogicalMatrix`. We use these types in a similar way to how we used `NumericVector`'s. The main differences are:

 * When we initialise, we need specify the number of rows and columns
    ```{Rcpp, eval=FALSE}
    // 10 rows, 5 columns
    NumericMatrix mat(10, 5);
    // Length 10
    NumericVector v(10);
    ```

 * We subset using `()`, i.e. `mat(5, 4)`.
 * The first view in a matrix is `mat(0, 0)` - remember indexes start with `0`.
 * To determine the number of rows and columns, we use the `.nrow()` and `.ncol()` methods.

# C++ with sugar on top

Rcpp sugar brings a higher-level of abstraction to C++ code written using the Rcpp API. What this means in practice is that we can write C++ code in the style of R. 

Let's suppose we want to create a C++ function that finds the squared difference between two R vectors; the squared residual in a regression analysis. Making sure we have the correct headers, we would try something like

```{Rcpp eval=FALSE}
NumericVector res_c(NumericVector x, NumericVector y){
  int i;
  int n = x.size();
  NumericVector residuals(n);
  for(i=0; i<n; i++) {
    residuals[i] = pow(x[i] - y[i], 2);
  }
  return residuals;
}
```

\noindent With `Rcpp` sugar we can rewrite this code to be more succinct and have more of an R feel, 

```{Rcpp eval=FALSE}
NumericVector res_sugar(NumericVector x, NumericVector y) {
  return pow(x-y, 2);
}
```

\noindent The sugar versions aren't usually faster than the C++ version, but there's usually very little difference between the two. However with the sugared variety, the code is shorter and is constantly being improved.

\subsection*{Exercises}

\begin{enumerate}
\item Construct an R version, \texttt{res\_r} and compare the three function variants.
\item In the above example, \texttt{res\_sugar} is faster than \texttt{res\_c}. Do you know why?
\end{enumerate}


