---
title: "Datasets"
output: pdf_document
---

\newpage

# Appendix: Datasets 

```{r, echo=FALSE}
# Plan: put a description of the datasets, and how to load them,
# in a chapter at the end
```

## Mobile Century travel behaviour dataset

This dataset is the result of an experiment conducted on the 8th February 2008,
10:00 to 18:00 (PST) on the Interstate 880 road in California.
100 GPS-enabled smart-phones were placed in cars for the experiment, the
aim of which was to evaluate the potential of smart-phones to be used to
monitor traffic conditions in real-time [@Herrera2010]. A website has been
set-up to describe and disseminate the data for research purposes:
[traffic.berkeley.edu/project/](http://traffic.berkeley.edu/project/).


```{r}
if(!file.exists("data/MobileCentury/pems_prop_NB.csv")){
  url1 <- "http://traffic.berkeley.edu/sites/default/files/downloads/MobileCentury_data_final_ver3.zip?sid=1529"
  url2 <- "http://traffic.berkeley.edu/sites/default/files/downloads/mobile_century_data_manual.pdf" # the user manual
  dir.create("data/MobileCentury") # create directory for the data
  downloader::download(url1, destfile = "data/MobileCentury_data_final_ver3.zip")
  downloader::download(url2, destfile = "data/MobileCentury/mobile_century_data_manual.pdf")
  unzip("data/MobileCentury_data_final_ver3.zip", exdir = "data/MobileCentury/")
}
```

## Moby Dick; or The Whale, by Herman Melville

The text file for the ebook of Moby Dick can be downloaded from the Gutenberg.org website\sidenote{\url{https://www.gutenberg.org/files/2701/2701.txt}}


