---
title: "Chapter 5 Visualisation"
author: "Colin Gillespie"
date: "17-18 September 2015"
output: ioslides_presentation
---

## 5.1 Introduction to ggplot2

  * `ggplot2` is a bit different from other graphics packages. 
  * It roughly follows the _philosophy_ of Wilkinson, 1999. 
  * Essentially, we think about plots as layers. 
  * By thinking of graphics in terms of layers it is
easier for the user to iteratively add new components and for a developer to add
new functionality.


## Example: the mpg data set

The `mpg` data set comes with the `ggplot2` package and can be using loaded in the usual way
```{r}
data(mpg, package="ggplot2")
```
This data set contains statistics on $234$ cars. 


## 5.1 Introduction to ggplot2

```{r, echo=T, message=FALSE, results="hide", fig.keep=FALSE}
plot(mpg$displ, mpg$cty, col=mpg$cyl)
```

[](figures/ch6_f1.png)



## 5.1 Introduction to ggplot2

After loading the necessary package

```{r message=FALSE}
library("ggplot2")
```

## 5.1 Introduction to ggplot2

```{r fig.keep='none', cache=TRUE, echo=T}
g = ggplot(data=mpg, aes(x=displ, y=cty))
g + geom_point(aes(colour=factor(cyl)))
```
[](figures/ch6_f2.png)

## 5.1 Introduction to ggplot2

  * The _ggplot_ function sets the default data set, and attributes called _aesthetics_. 
  * The aesthetics are properties that are perceived on the graphic. 
  * A particular aesthetic can be mapped to a variable or set to a constant
value. 
  * In previous figure, the variable `displ` is mapped to the x-axis and `cty` variable is mapped to the y-axis. 


## 5.1 Introduction to ggplot2

  * The other function, `geom_point` adds a layer to the plot. 
  * The `x` and `y` variables are inherited (in this case) from the first function, `ggplot`, and the colour aesthetic is set to the `cyl` variable. 
  * Other possible aesthetics are, for example, size, shape and transparency. 

## 5.1 Introduction to ggplot2

If instead we changed the `size` aesthetic

```{r cache=TRUE, echo=TRUE}
g + geom_point(aes(size=factor(cyl)))
```


## 5.1 Introduction to ggplot2

Plot Name | Geom  | Base graphic 
----------|-------|-----------------
Barchart | bar  | __barplot__
Box-and-whisker | boxplot | __boxplot__
Histogram | histogram  | __hist__
Line plot | line | __plot__ and __lines__
Scatter plot | point | __plot__ and __points__

# The bigvis package
## The bigvis package

  * The `bigvis` package provides tools for exploratory data analysis of large datasets ($10-100$ million obs).
  * The goal is that operations should take less than $5$ seconds on a standard computer, even when the sample size is $100$ million. 
  * The package is currently not available on CRAN

```{r eval=FALSE, tidy=FALSE}
devtools::install_github("hadley/bigvis")
```

  * If you are using Windows, you will also need to install Rtools. 


## The bigvis package

  * Directly visualising raw big data is pointless. 
  * It's a waste of time to create a $100$ million point scatter plot, since we would not be able to distinguish between the points. 
  * In fact, we  are likely to run out of pixels! If you doubt this, compare these two plots

```{r fig.keep="none"}
par(mfrow=c(1, 2))
plot(1, 1,ylab="")
plot(rep(1, 1e3), rep(1, 1e3), ylab="")
```
  * Instead, we need to quickly summarise the data and plot the data in a sensible way.


## The bigvis package

  * Similar to `dplyr`
  * It provides fast C++ functions to manipulate the data, with the resulting output being handled by standard R functions (but optimised for `ggplot2`). 
  * The package also provides a few functions for handling outliers, since when visualising big data outliers may be more of an issue.


## Bin and condense

  * The `bin()` and `condense()` functions are used to get compact summaries of the data.     * For example, suppose we generate $10^5$ random numbers from the $t$ distribution
    ```{r echo=2}
    set.seed(1)
    x = rt(1e5, 5)
    ```


## Bin and condense

  * The `bin` and `condense` functions create the  binned variable
    ```{r message=FALSE}
    library("bigvis")
    ## Bin in blocks of 0.01
    x_sum = condense(bin(x, 0.01))
    ```
  * After binning you may want to smooth out any rough estimates (similar to kernel density estimation). 

## Smooth

```{r echo=1:2}
## h is the binwidth (similar to bin size)
x_smu = smooth(x_sum, h = 5 / 100)
par(mar=c(3,3,2,1), mgp=c(2,0.4,0), tck=-.01,
                      cex.axis=0.9, las=1)
plot(x_sum, panel.first=grid(), xlim=c(-12, 12), 
     ylab="Count", pch=21, cex=0.6)
lines(x_smu, col=2, lwd=2)
text(5, 200, "Smoothed line", col=2)
```

## Autoplot

```{r}
autoplot(x_sum)
```

## Autoplot + peel

```{r}
autoplot(peel(x_smu))
```

<!-- From Alastair Sanderson -->

## IMDB example

```{r}
data(movies, package="bigvis")
```

The dataset is a data frame and has `r NCOL(movies)` columns and `r NROW(movies)` rows. 

## IMDB example

We create bin versions of the movie length and rating using the `condense/bin` trick

```{r tidy=FALSE, message=FALSE}
n_bins = 1e4
bin_data = with(movies, 
    condense(bin(length, find_width(length, n_bins)),
             bin(rating, find_width(rating, n_bins))))
```

## IMDB example

```{r echo=1}
ggplot(bin_data, aes(length, rating, fill=.count )) + 
  geom_raster()
```

## IMDB example

The resulting plot isn't helpful, due to a couple of long movies

```{r tidy=FALSE}
## Longer than one day!!
subset(movies[ ,c("title", "length", "rating")], 
       length > 24*60)
```

## IMDB example: last_plot + peel

```{r echo=1, fig.keep="last"}
last_plot() %+% peel(bin_data)
ggplot(data=peel(bin_data), aes(length, rating, fill=.count )) + 
  geom_raster()
```

\noindent to get a better visualisation.

## Tableplots: the tabplot package

  * Tableplots are a visualisation technique that can be used to explore and analyse large data sets. 
  * These plots can be used to explore variable relationships and check data quality.   
  * Tableplots can visualise multivariate datasets with several variables and a large number of records. 
  * The `tabplot` package provides has an `ffdf` interface.


## Tableplots: the tabplot package

```{r, echo=T,  message=FALSE}
library("tabplot")
tableplot(movies[,3:5])
```


## Tableplots: the tabplot package

```{r message=FALSE, warning=FALSE}
tableplot(movies[,3:5], sortCol = 3)
```


## Tableplots: the tabplot package

```{r message=FALSE, warning=FALSE, tidy=FALSE}
tableplot(movies[,3:5], sortCol = 3, from =0, to=10)
```
