---
title: "Memory"
author: "Colin Gillespie"
date: "17-18 September 2015"
output: ioslides_presentation
---

# Chapter 5: Visualisation

## 5.1 Introduction to ggplot2

  * `ggplot2` is a bit different from other graphics packages. 
  * It roughly follows
the \textit{philosophy} of Wilkinson, 1999. 
  * Essentially, we
think about plots as layers. 
  * By thinking of graphics in terms of layers it is
easier for the user to iteratively add new components and for a developer to add
new functionality.


## Example: the mpg data set

The `mpg` data set comes with the `ggplot2` package and can be using loaded in the usual way
```{r}
data(mpg, package="ggplot2")
```
This data set contains statistics on $234$ cars. 


## 5.1 Introduction to ggplot2

```{r, echo=T, message=FALSE, results="hide", fig.keep=FALSE}
plot(mpg$displ, mpg$cty, col=mpg$cyl)
```

[](figures/ch6_f1.png)



## 5.1 Introduction to ggplot2

After loading the necessary package

```{r message=FALSE}
library("ggplot2")
```

## 5.1 Introduction to ggplot2

```{r fig.keep='none', cache=TRUE, echo=T}
g = ggplot(data=mpg, aes(x=displ, y=cty))
g + geom_point(aes(colour=factor(cyl)))
```
[](figures/ch6_f2.png)

## 5.1 Introduction to ggplot2

  * The
\texttt{ggplot} function sets the default data set, and attributes called
\textbf{aesthetics}. 
  * The aesthetics are properties that are perceived on the
graphic. 
  * A particular aesthetic can be mapped to a variable or set to a constant
value. 
  * In figure \ref{F6.2}, the variable \texttt{displ} is mapped to the x-axis and
\texttt{cty} variable is mapped to the y-axis. 


## 5.1 Introduction to ggplot2

The other function, \texttt{geom\_point} adds a layer to the plot. The \texttt{x} and
\texttt{y} variables are inherited (in this case) from the first function, \texttt{ggplot}, and
the colour aesthetic is set to the \texttt{cyl} variable. Other possible aesthetics
are, for example, size, shape and transparency. In figure \ref{F6.2} these
additional aesthetics are left at their default value.

If instead we changed the `size` aesthetic

```{r cache=TRUE, echo=2}
png("figures/ch6_3.png", width=5*dpi, height=5*dpi, res=dpi)
g + geom_point(aes(size=factor(cyl)))
sink=dev.off()
```

\noindent we would get figure \ref{F6.3} where the size of the points vary with `cyl`. Table \ref{T6.1} gives a summary of standard geoms.


\begin{table}[t]
  \centering
  \begin{tabular}{@{}lll@{}}
    \toprule
    Plot Name & Geom  & Base graphic \\
    \midrule
    Barchart & bar  & \texttt{barplot}\\
    Box-and-whisker & boxplot & \texttt{boxplot}\\
    Histogram & histogram  & \texttt{hist} \\
    Line plot & line & \texttt{plot} and \texttt{lines}\\
    Scatter plot & point & \texttt{plot} and \texttt{points}\\
    \bottomrule
  \end{tabular}
  \caption[4\baselineskip]{Basic \texttt{geom}'s and their corresponding standard plot names.}\label{T6.1}
\end{table}


\begin{marginfigure}
  \centering
  \includegraphics[]{figures/ch6_3.png}
  \caption{As figure \ref{F6.2}, but where the size aesthetic depends on
    cylinder size.}\label{F6.3}
\end{marginfigure}
  
# The bigvis package

The `bigvis` package provides tools for exploratory data analysis of large datasets ($10-100$ million obs).
The goal is that operations should take less than $5$ seconds on a standard computer, even when the sample size is $100$ million. The package is currently not available on CRAN and needs to be installed directly from github using the `devtools` package

```{r eval=FALSE, tidy=FALSE}
devtools::install_github("hadley/bigvis")
```

\noindent If you are using Windows, you will also need to install Rtools. 

Directly visualising raw big data is pointless. It's a waste of time to create a $100$ million point scatter plot, since we would not be able to distinguish between the points. In fact, we  are likely to run out of pixels! If you doubt this, compare these two plots

```{r fig.keep="none"}
par(mfrow=c(1, 2))
plot(1, 1,ylab="")
plot(rep(1, 1e3), rep(1, 1e3), ylab="")
```

\noindent Except for some anti-aliasing issues, it's impossible to tell the difference between these two plots. Instead, we need to quickly summarise the data and plot the data in a sensible way.

Similar to `dplyr`, the `bigvis` package is structured around a few key functions. It provides fast C++ functions to manipulate the data, with the resulting output being handled by standard R functions (but optimised for `ggplot2`). The package also provides a few functions for handling outliers, since when visualising big data outliers may be more of an issue.


\subsection*{Bin and condense}

The `bin()` and `condense()` functions are used to get compact summaries of the data. For example, suppose we generate $10^5$ random numbers from the $t$ distribution
```{r echo=2}
set.seed(1)
x = rt(1e5, 5)
```

\noindent The `bin` and `condense` functions create the  binned variable
```{r message=FALSE}
library("bigvis")
## Bin in blocks of 0.01
x_sum = condense(bin(x, 0.01))
```

\subsection*{Smooth}

After binning you may want to smooth out any rough estimates (similar to kernel density estimation). The `smooth` function smooths out the binned data 

```{r echo=1:2}
## h is the binwidth (similar to bin size)
x_smu = smooth(x_sum, h = 5 / 100)
png("figures/ch6_4.png", width=5*dpi, height=5*dpi, res=dpi)
par(mar=c(3,3,2,1), mgp=c(2,0.4,0), tck=-.01,
                      cex.axis=0.9, las=1)
plot(x_sum, panel.first=grid(), xlim=c(-12, 12), 
     ylab="Count", pch=21, cex=0.6)
lines(x_smu, col=2, lwd=2)
text(5, 200, "Smoothed line", col=2)
sink=dev.off()
```

\begin{marginfigure}
  \centering
  \includegraphics[]{figures/ch6_4.png}
  \caption{Black points are the binned data. Red line is the smoothed estimate.}\label{F6.4}
\end{marginfigure}

\noindent Consult the functions `best_h()` and `rmse_cvs()` to get an idea of a good starting binwidth.

\subsection*{Visualisation}

The output of the the `condense` and `smooth` functions can be visualised using standard plotting packages. The `bigvis` package also contains an `autoplot` function to quickly visualise results

\begin{marginfigure}
  \centering
  \includegraphics[]{figures/ch6_5.png}
  \includegraphics[]{figures/ch6_6.png}
  \caption{Plots generated using `autoplot`. The bottom graph is the `peeled` version of the data.}\label{F6.5}
\end{marginfigure}

```{r echo=2}
png("figures/ch6_5.png", width=5*dpi, height=5*dpi, res=dpi)
autoplot(x_sum)
sink=dev.off()
```

\noindent This can be combined with the handy `peel` function, that (by default) just contains the middle 99% of the data

```{r echo=2}
png("figures/ch6_6.png", width=5*dpi, height=5*dpi, res=dpi)
autoplot(peel(x_smu))
sink=dev.off()
```

<!-- From Alastair Sanderson -->

## IMDB example

The internet movie database (IMDB)\sidenote{\url{http://imdb.com/}} is a website devoted to collecting movie data supplied by studios and fans. It claims to be the biggest movie database on the web and is run by Amazon. A version of the data set comes with the `bigvis` package

```{r}
data(movies, package="bigvis")
```

\noindent The dataset is a data frame and has `r NCOL(movies)` columns and `r NROW(movies)` rows. We create bin versions of the movie length and rating using the `condense/bin` trick

```{r tidy=FALSE, message=FALSE}
n_bins = 1e4
bin_data = with(movies, 
    condense(bin(length, find_width(length, n_bins)),
             bin(rating, find_width(rating, n_bins))))
```

\noindent This data set can then plotted as a heatmap using

```{r echo=2}
png("figures/ch6_7.png", width=5*dpi, height=5*dpi, res=dpi)
ggplot(bin_data, aes(length, rating, fill=.count )) + 
  geom_raster()
sink=dev.off()
```
\begin{figure}[t]
  \centering
  \includegraphics[width=0.5\textwidth]{figures/ch6_7.png}%
  \includegraphics[width=0.5\textwidth]{figures/ch6_8.png}
  \caption{Movie Rating vs Length. The plot on the right is the peeled version.}\label{F6.6}
\end{figure}

\noindent The resulting plot isn't helpful, due to a couple of long movies

```{r tidy=FALSE}
## Longer than one day!!
subset(movies[ ,c("title", "length", "rating")], 
       length > 24*60)
```

\noindent The `ggplot2` package contains a handy function called `last_plot` that allows us to manipulate the last created plot. For this example, we'll manipulate the plot using the `peel` function

```{r echo=1, fig.keep="none"}
last_plot() %+% peel(bin_data)
png("figures/ch6_8.png", width=5*dpi, height=5*dpi, res=dpi)
ggplot(data=peel(bin_data), aes(length, rating, fill=.count )) + 
  geom_raster()
sink=dev.off()
```

\noindent to get a better visualisation. The associated paper
\begin{center}
\url{http://vita.had.co.nz/papers/bigvis.pdf}
\end{center}
\noindent provides a good introduction to the key ideas.

# Tableplots: the tabplot package

Tableplots are a visualisation technique that can be used to explore and analyse large data sets. These plots can be used to explore variable relationships and check data quality. Tableplots can visualise multivariate datasets with several variables and a large number of records. The `tabplot` package provides has an `ffdf` interface.
\begin{marginfigure}
  \centering
  \includegraphics[]{figures/ch6_9.png}
  \caption{Tableplot of the movie dataset.}\label{F6.5}
\end{marginfigure}

In a tableplot, numeric variables are plotted as histograms of the mean values while for categorical variable, stacked bar charts are used to show category proportions. Missing values are also highlighted.

Since `tableplot` can not handle character columns, when plotting we'll remove the first column and for presentation just select three columns


```{r, echo=c(1, 3), fig.keep="none", message=FALSE}
library("tabplot")
png("figures/ch6_9.png", width=5*dpi, height=5*dpi, res=dpi)
tableplot(movies[,3:5])
sink=dev.off()
```

\noindent By default, the first column is sorted, but this can be altered using the `sortCol` argument

```{r fig.keep="none", message=FALSE, warning=FALSE}
tableplot(movies[,3:5], sortCol = 3)
```

\noindent It is also possible to zoom into key sections of the plot. For example, if we wanted to zoom into the top 10\% of movies according to rating, then we can use the `from` and `to` arguments. 
\begin{marginfigure}
  \centering
  \includegraphics[]{figures/ch6_10.png}
  \caption{Tableplot of the movie dataset. Only the top 10\% of movies (based on rating) have been plotted.}\label{F6.8}
\end{marginfigure}

```{r fig.keep="none", message=FALSE, warning=FALSE, tidy=FALSE, echo=2}
png("figures/ch6_10.png", width=5*dpi, height=5*dpi, res=dpi)
tableplot(movies[,3:5], sortCol = 3, from =0, to=10)
sink=dev.off()
```

\noindent For a detailed description of the package consult the vignette

```{r eval=FALSE}
browseVignettes("tabplot")
```

# Interactive visualisations

One of the features of large, complex datasets is that they can be visualised in many ways.
Sometimes it is only by viewing the relationships between different variables *interactively* that patterns in the data become apparent.
Interactive visualisation is also a way to communicate the contents of large datasets to others, without needing to transfer huge files from one computer to another.
Perhaps the main advantage of online interactive visualisation, however, is that the 'heavy lifting' to produce the plots can be done remotely.
If powerful computers to handle the processing 'server side', the user's computer is freed from the strain that large datasets can put on hard-discs, RAM and CPU.

Because of these advantages interactive visualisation has become increasingly common amongst R users over the last few years.
It is a rapidly evolving field within R.
Instead of providing code examples, this section therefore highlights some promising packages and provides concrete examples with reference to additional teaching material, to show what is possible

## Shiny

**shiny** is an R package for easing the development of online, interactive web applications ('apps').
An example of a **shiny** app for real world use is the Propensity to Cycle Tool (PCT), funded by the UK's Department for Transport (figure 6.10).
It is necessary to host the visualisation remotely because the target audience (local transport planners) cannot be expected to download all the input data and software dependencies to run the model locally.
Moreover, most policy makers and not R experts so the visualisation reduces the barriers to entry into exploratory data analysis.

```{r cov, fig.margin=TRUE, fig.cap= "Screenshot of the Propensity to Cycle Tool shiny app", echo=FALSE}
grid.raster(readPNG("figures/coventry-centroids.png"))
```

There are many excellent **shiny** teaching resources, the best of which is RStudio's website dedicated to shiny:
[shiny.rstudio.com/](http://shiny.rstudio.com/).

## ggvis

**ggvis** builds on **ggplot2** to ease the creation of interactive plots that be used locally or pushed online for use by others.
A user-friendly tutorial teaching **ggvis** basics and showcasing its capabilities can be found online, at
[ggvis.rstudio.com/ggvis-basics.html](http://ggvis.rstudio.com/ggvis-basics.html).

## taucharts

**taucharts** is a new package for creating interactive graphics using the JavaScript library TauCharts.
The package has a syntax that is similar to **ggplot2** but provides options for user-interaction.
Read more about the package at [rpubs.com/hrbrmstr/taucharts](http://rpubs.com/hrbrmstr/taucharts).
