---
title: "dplyr"
author: "Robin Lovelace"
date: "September 18, 2015"
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = "../")
```

## Where we got to last time

```{r, message=FALSE}
pkgs <- c("tidyr", "readr", "dplyr")
pkld <- lapply(pkgs, require, character.only = TRUE)
raw <- read_csv("data/pew.csv")
rawt <- gather(raw, Income, Count, -religion)
head(rawt, 3)
```

## Filtering columns

```{r}
df <- read_csv("data/miniaa") # load imaginary large data
dim(df)
all_na <- sapply(df, function(x) all(is.na(x)))
df1 <- df[!all_na] # subset the dataframe
dim(df1)
df2 <- df[complete.cases(t(df))]
dim(df2)
```

## Aggregation

```{r, warning=FALSE}
df <- read_csv("data/ghg-ems.csv")
glimpse(df)
```

## What to aggregate?

```{r}
df <- rename(df, ECO2 = `Electricity/Heat (CO2) (MtCO2)`)
length(unique(df$Country))
sapply(df, function(x) length(unique(x)))
```

## Aggregation with base R

```{r}
e_ems <- aggregate(df$ECO2, list(df$Country), mean, na.rm = T)
nrow(e_ems)
head(e_ems) # not particularly beautiful output
```

## Making base R code nicer

```{r}
e_ems <- aggregate(ECO2 ~ Country, df, mean, na.rm = T)
head(e_ems)
```

## Enter dplyr

```{r}
e_ems <- group_by(df, Country) %>%
  summarise(mean_eco2 = mean(ECO2, na.rm = T))
e_ems
```

## dplyr: data processing made easy

```{r}
glimpse(e_ems)
```

# Practical: work through chapter 5.3 and 5.4 (15 minutes)

## First thing to do

> - Rename the 4th column of the data
> - Optional challenge: do this in the dplyr way
> - Advanced: Find the top and bottom 3 emitters in each category

```{r}
df <- read_csv("data/ghg-ems.csv")
names(df)
names(df)[4] <- "ECO2"
```

## Command chaining I

Which do you prefer:

> - This?

```{r, eval=FALSE}
top_n(
  arrange(
    summarise(
      group_by(
        filter(idata, grepl("g", Country)),
        Year),
      gini = mean(gini, na.rm = T)),
    desc(gini)),
  n = 5)
```


## Command chaining II

> - Or this?

```{r, eval=FALSE}
idata %>%
  filter(grepl("g", Country)) %>%
  group_by(Year) %>%
  summarise(gini = mean(gini, na.rm = T)) %>%
  arrange(desc(gini)) %>%
  top_n(n = 5)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.

# Practical: Section 5.5 (10 minutes)
